A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-22 tidyindex.qmd
rework on the writing of GGGI  example

A linear combination can also be interpreted as a linear projection of multivariate information with a weight vector. Projecting data from higher to lower dimensions unavoidably leads to information loss and the weights used require careful examination to understand their effects on the index and implications for interpretation and decision-making. By making slight adjustments to the weight vector, we can observe how index values and country rankings change. For illustration, we select a set of countries including:

1)  Top-ranked countries with GGGI \> 0.85,
2)  Countries ranked between 57 and 62, with GGGI values from 0.72 to 0.73, and
3)  Low-ranked countries with GGGI \< 0.6.

We slightly vary the weight of the politics dimension from the original 0.25 while keeping the weights constant for the other three dimensions. This process creates an animation showing the movement of index values in response to changing weights. This visualization technique, which presents a sequence of data projections, is referred to as a "tour". The specific kind of tour that moves between the original and nearby projections is known as a "local tour". XXX-ref to tourr package, and also tour theory paper (Buja https://doi.org/10.1016/S0169-7161(04)24014-7 )

In @fig-idx-tour, six frames have been chosen from the animation available at https://vimeo.com/847874016. When the weight of politics is reduced (Frame 1 and 6 vs. Frame 12), the difference in GGGI values between the top Nordic countries and mid- or low-ranked countries narrows, suggesting a smaller variation among countries in achieving gender parity. Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs. Frame 12), nearly all the countries in the three categories experience a decrease in GGGI values. A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight. This leads to its index value being almost similar to those of the top-ranked Nordic countries when the politics dimension is assigned a weight of 0.52 in Frame 29.

In GGGI, the index value has a direct interpretation as the percentage of the gender gap that has been closed. When countries exhibit closer or consistent decreasing index values, it can be directly interpreted as a reflection of a country's progress toward gender parity. Ideally, an index should be robust against minor changes in its construction components. This example provides researchers with means to observe changes in the index resulting from variations in the weights used in linear combinations. This approach can be applied broadly to different sets of weights of interest, extending beyond the change in a single dimension illustrated by the example.

>>>>

Frame 12 in @fig-idx-tour presents the original index values when equal weights are applied across the four dimensions. Leading the rankings, Nordic countries (Iceland, Norway, and Finland) and New Zealand show a good progression on reaching gender parity, with over 80% of the gender gap closed. In the middle range, an index value over 0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan, Armenia, and Slovakia. The last group of countries (Pakistan, Iran, Algeria, Chad), with GGGI < 0.6, have a wider gender gap to closed and Afghanistan lags further behind with GGGI = 0.4.

To observe the effect of a minor weight change on the index, we slightly vary the politics weight (while maintain equal weights among other dimensions) to generate an animation of the movement of index values in response to the changing weights. This visualization technique, which presents a sequence of data projections, is referred to as a "tour" and the specific kind of tour used here to move between nearby projections is known as a "local tour" [@buja2005computational; @wickham_tourr_2011].

In @fig-idx-tour, six frames have been chosen from the local tour to explore the role of the politics dimension to the index. When the weight of politics is reduced (Frame 1 and 6 vs. Frame 12), the difference in GGGI values between the Nordic and other countries narrows, suggesting a smaller variation among countries in achieving gender parity. Conversely, when the weight of politics increases (Frame 18, 24, and 29 vs. Frame 12), nearly all the countries experience a decrease in GGGI values. A noticeable exception to this trend is Bangladesh, where its index value moves in the same direction as the politics weight, which leads to an almost similar index value to the Nordics when the politics dimension is assigned a weight of 0.52 in Frame 29.

Ideally, an index should be robust against minor changes in its construction components. Unfortunately, this is not the case with GGGI, which raises concerns about the index's quality. For other indexes that use linear combination to reduce variable dimensions, the demonstrated local tour can be applied similarly to assess the index robustness through observing how index values respond to perturbation of the weights.

### 2023-08-22 tidyindex.tex
rework on the writing of GGGI  example

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, sharp corners, breakable, enhanced, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Examples of using a pipeline approach for data analysis can be traced
back to the interactive graphics literature, including Buja et al.
(1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
Wickham et al. (2009). Wickham et al. (2009) argue that whether made
explicit or not, a pipeline has to be presented in every graphics
program, and making them explicit is beneficial for understanding the
implementation and comparing between different graphic systems. While
this comment is made in the context of interactive graphics programs, it
is also applicable generally to any data analysis workflow. More
recently, the tidyverse suite (Wickham et al. 2019) takes the pipeline
approach for general-purpose data wrangling and has gained popularity
within the R community. The pipeline-style code can be directly read as
a series of operations applied successively on tidy data objects,
offering a method to document the data wrangling process with all the
computational details for reproducibility.

@@

A linear combination can also be interpreted as a linear projection of
multivariate information with a weight vector. Projecting data from
higher to lower dimensions unavoidably leads to information loss and the
weights used require careful examination to understand their effects on
the index and implications for interpretation and decision-making. By
making slight adjustments to the weight vector, we can observe how index
values and country rankings change. For illustration, we select a set of
countries including:

\begin{enumerate}
\def\labelenumi{\arabic{enumi})}
\tightlist
\item
  Top-ranked countries with GGGI \textgreater{} 0.85,
\item
  Countries ranked between 57 and 62, with GGGI values from 0.72 to
  0.73, and
\item
  Low-ranked countries with GGGI \textless{} 0.6.
\end{enumerate}

We slightly vary the weight of the politics dimension from the original
0.25 while keeping the weights constant for the other three dimensions.
This process creates an animation showing the movement of index values
in response to changing weights. This visualization technique, which
presents a sequence of data projections, is referred to as a ``tour''.
The specific kind of tour that moves between the original and nearby
projections is known as a ``local tour''. XXX-ref to tourr package, and
also tour theory paper (Buja
https://doi.org/10.1016/S0169-7161(04)24014-7 )

In Figure~\ref{fig-idx-tour}, six frames have been chosen from the
animation available at https://vimeo.com/847874016. When the weight of
politics is reduced (Frame 1 and 6 vs.~Frame 12), the difference in GGGI
values between the top Nordic countries and mid- or low-ranked countries
narrows, suggesting a smaller variation among countries in achieving
gender parity. Conversely, when the weight of politics increases (Frame
18, 24, and 29 vs.~Frame 12), nearly all the countries in the three
categories experience a decrease in GGGI values. A noticeable exception
to this trend is Bangladesh, where its index value moves in the same
direction as the politics weight. This leads to its index value being
almost similar to those of the top-ranked Nordic countries when the
politics dimension is assigned a weight of 0.52 in Frame 29.

In GGGI, the index value has a direct interpretation as the percentage
of the gender gap that has been closed. When countries exhibit closer or
consistent decreasing index values, it can be directly interpreted as a
reflection of a country's progress toward gender parity. Ideally, an
index should be robust against minor changes in its construction
components. This example provides researchers with means to observe
changes in the index resulting from variations in the weights used in
linear combinations. This approach can be applied broadly to different
sets of weights of interest, extending beyond the change in a single
dimension illustrated by the example.

@@

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, interior hidden, boxrule=0pt, enhanced, breakable, sharp corners]}{\end{tcolorbox}}\fi

@@

Examples of using a pipeline approach for data analysis can be traced
back to the interactive graphics literature, including A. Buja et al.
(1988); Sutherland et al. (2000); Xie, Hofmann, and Cheng (2014);
Wickham et al. (2009). Wickham et al. (2009) argue that whether made
explicit or not, a pipeline has to be presented in every graphics
program, and making them explicit is beneficial for understanding the
implementation and comparing between different graphic systems. While
this comment is made in the context of interactive graphics programs, it
is also applicable generally to any data analysis workflow. More
recently, the tidyverse suite (Wickham et al. 2019) takes the pipeline
approach for general-purpose data wrangling and has gained popularity
within the R community. The pipeline-style code can be directly read as
a series of operations applied successively on tidy data objects,
offering a method to document the data wrangling process with all the
computational details for reproducibility.

@@

Frame 12 in Figure~\ref{fig-idx-tour} presents the original index values
when equal weights are applied across the four dimensions. Leading the
rankings, Nordic countries (Iceland, Norway, and Finland) and New
Zealand show a good progression on reaching gender parity, with over
80\% of the gender gap closed. In the middle range, an index value over
0.7 is achieved for Brazil, Panama, Poland, Bangladesh, Kazakhstan,
Armenia, and Slovakia. The last group of countries (Pakistan, Iran,
Algeria, Chad), with GGGI \textless{} 0.6, have a wider gender gap to
closed and Afghanistan lags further behind with GGGI = 0.4.

To observe the effect of a minor weight change on the index, we slightly
vary the politics weight (while maintain equal weights among other
dimensions) to generate an animation of the movement of index values in
response to the changing weights. This visualization technique, which
presents a sequence of data projections, is referred to as a ``tour''
and the specific kind of tour used here to move between nearby
projections is known as a ``local tour'' (Andreas Buja et al. 2005;
Wickham et al. 2011).

In Figure~\ref{fig-idx-tour}, six frames have been chosen from the local
tour to explore the role of the politics dimension to the index. When
the weight of politics is reduced (Frame 1 and 6 vs.~Frame 12), the
difference in GGGI values between the Nordic and other countries
narrows, suggesting a smaller variation among countries in achieving
gender parity. Conversely, when the weight of politics increases (Frame
18, 24, and 29 vs.~Frame 12), nearly all the countries experience a
decrease in GGGI values. A noticeable exception to this trend is
Bangladesh, where its index value moves in the same direction as the
politics weight, which leads to an almost similar index value to the
Nordics when the politics dimension is assigned a weight of 0.52 in
Frame 29.

Ideally, an index should be robust against minor changes in its
construction components. Unfortunately, this is not the case with GGGI,
which raises concerns about the index's quality. For other indexes that
use linear combination to reduce variable dimensions, the demonstrated
local tour can be applied similarly to assess the index robustness
through observing how index values respond to perturbation of the
weights.

@@

\leavevmode\vadjust pre{\hypertarget{ref-buja2005computational}{}}%
Buja, Andreas, Dianne Cook, Daniel Asimov, and Catherine Hurley. 2005.
{``Computational Methods for High-Dimensional Rotations in Data
Visualization.''} \emph{Handbook of Statistics} 24: 391--413.
\url{https://doi.org/10.1016/S0169-7161(04)24014-7}.
